---
title: "jigsaw lookat"
author: "Amy Goldlist"
date: "24/02/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Jigsaw Puzzle Problem

Lately, I've been playing a lot of jigsaw puzzles on my phone.  One day, I noticed that when I started assembling the puzzle, focusing on the edges, I usually see two matching pieces next to each other.  What are the odds? I marveled.

And so, I obviously needed to calculate that.
First, I ran a simple simulation, to check some experimental results.  My favourite difficulty level of puzzle is a 15 by 15 puzzle which contains 225 pieces, of which 58 are edge pieces (feel free to count!).  I started by asking:
If I reorder the numbers between 1 and n, what is the probability that my sequence contains at least one consecutive pair?  (Here $n$ = 58)
I ran this with different $n$ values, and discovered that once $n$ got big enough (say 10 or more), it seemed to always give and answer between 85% and 90%.  Interesting.  But it must depend on $n$, yes?

For the sake of ease, I decided to not worry about the fact that my edges actually form a chain â that is, piece 1 is next to piece 2.  I'll correct that at the end.  [Note to Amy: Correct].

First,
Note:
* There are $n!$ possible orderings of $n$ objects.
* There are $n$ consecutive pairs [note $n-1$ if 1 isn't next to $n$]
* There are $n \choose 2$ (n choose 2) possible pairs,

So:
$$\text{Number of possible pairs} = \frac{n!}{2(n-2)!}= \frac{n(n-1)}{2}$$
So, if I choose 2 pieces at random, the odds of them being consecutive are:

\[P(match)= \frac{\text{#  Consecutive pairs}}{\text{total possible pairs}} = \frac{n}{\left(\frac{n(n-1)}{2}\right)} = \frac{2}{n-1}\]


Now, back to the puzzle:
We are working in a binomial distribution with:
$n-1$ as the number of trials (that is, the number of possible pairs in my sequence)
$\pi= \frac{2}{n-1}$ is the probability of success:
And I am looking for $P(X \geq 0) = 1- P(X =0)$:

So:
\[1-(n-1)C0 \times \frac{2}{n-1}^0 \left(1-\frac{2}{n-1}\right)^{n-1}\]
\[ =  \left(1-\frac{2}{n-1}\right)^{n-1}\]
For n = 58, we have:
\[P = 1- \left(1-\frac{2}{57}\right)^{57} = 0.8694411\]

```{r}
1-(1-2/57)^57

```

But what if $n$ is big?
Now we have
\[lim_{n \rightarrow \infty} = 1- \left(1-\frac{2}{n-1}\right)^{n-1} = 1-e^{-2}=0.8646647\]
```{r}
1-exp(-2)

```

Around 87%.





### Part 2:  The whole puzzle

As I proudly presented this to my family at dinner time, my kid wisely asked: "What about the whole puzzle?  Not just the edges?  Puzzle pieces touch 4 others - so would you multiply the answer by 4?"

Good question kid!  No, but as it turns out - close!

Let's start with an $n$ by $m$ puzzle.  This puzzle has $n(m-1)+m(n-1)$ possible matches.  Confused? Look at this nice image:

![image]

So we have the following numbers:

* $n$ columns and $m$ rows
* $m\times n = mn$ total pieces
* When we display the pieces in a rwo, there are $nm-1$ possible pairs.
* There are $n(m-1)+m(n-1) = 2mn-n-m$ total pairings of pieces that fit together.
* There are $mn \choose 2 = \frac{mn(mn-1)}{2}$ possible pairings.

That means that if I select 2 pieces at random, the probability that they are pairs (ie, they fit together) is:

\[P(fit)= p = \frac{2(2mn-n-m)}{mn(mn-1)}\]

So let's fit that into our same binomial distribution:

\[P(X>0)=$P(X \geq 0) = 1- P(X =0)\]

So:
\[1- ({(mn-1)\choose {0} }\times \frac{2(2mn-n-m)}{mn(mn-1)}^0 \left(1-\frac{2(2mn-n-m)}{mn(mn-1)}\right)^{mn-1}=1-\left(1-\frac{2(2mn-n-m)}{mn(mn-1)}\right)^{mn-1}\]

For my 15 by 15 puzzle, (ie, $m=n=15$), we have:

So:
\[1- \left(1-\frac{2(2mn-n-m)}{mn(mn-1)}\right)^{mn-1} =\left(1-\frac{2(420)}{225(224)}\right)^{224} = 0.9768276\]



```{r}
1-(1-(2*(2*225-30))/(225*224))^(224)
```
But I like to generalize!  Let's say this is a really big puzzle.  For simplicity's sake, let's make it an $n$ by $n$ square.  As $n$ grows large, we have:

\[lim_{n \rightarrow \infty}  1- \left(1-\frac{4}{n(n-1)}\right)^{n^2-1} = 1-e^{-4}=0.8646647\]

```{r}
1-exp(-4)
```
So in this case, I really would expect to see at least one match!